---
title: "Analytics 502"
author: "Avik Roy"
date: "September 30, 2018"
output:
  word_document: default
  pdf_document: default
  html_document: default
classoption: landscape
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=30),tidy=TRUE)
library(tidyverse)
library(corrplot)
set.seed(111)
```

# ABSTRACT
NOTE TO TEAM: So this should be a complete version of the research. I have focussed on this research on passengers. And passengers care about Arrival Delay most. And they can in most cases only have airline, date and time to 'pick. 

A quick summary below (not in presenatable prose, but for the teams info):

Goal: 
1.) Finding out what factors influence the delay of a flight. 
2.) Enabling a passenger to find when they should book a flight 
3.)Let passenger predict what can be their estimated arrival delay.

EDA: 
-> A few snippets to get a feel for the data
-> Scatterplot vs time should give a general idea of how flights delays trend over the week
-> Correlation plot checks the realtionship between the 'minutes' based caloums and distance. The focus should be ARRIVAL_DELAY. Specifically that it does not have much relatioship with other variables like distance. Other conclusions can be brought forth as well. Feel free.
-> Histogram plot **hints** that there is something really going on about departure day/time and the delay in causes (HINT: candidate for hypothesis testing)

INFERENCE:
-> Hypothesis test: As ofcourse is possible, we can do many many hypothesis tests. But for showing to prof that we CAN do it. I focussed on Saturday evening. Highest, and looks like a passenger should definitely avoid. So to make sure that I am not seeing a random manifestation of sampling and the mean delay is not in reality, '0', we conducted a hypothesis test.
-> ANOVA shows that the histogram above 'has significance'. And there is significant difference between the means of the day/time groups

THE 'BEST' MODEL:
Should be straight forward to understand if you have read the notes.
BE ADVISED: Adjusted R-square is very low. Which is FINE! The reason I say is the example shared in the class notes was also low. And this is real data, so there are probably a quadrillion other factors which make up the variance. Please read this blog article to learn how to interpret the regression results:http://blog.minitab.com/blog/adventures-in-statistics-2/how-to-interpret-a-regression-model-with-low-r-squared-and-low-p-values

PREDICTION:
Self explanatory

APPENDIX:
meh.....do what you will :)

NOTE TO SELF:
Clean this up for submitting to prof.
#DATA LOADING & MANIPULATION
####Checking out the data

```{r check}
flightData <- read.csv("flights.csv", header = TRUE)
head(flightData)
str(flightData)
summary(flightData)
```

####Modifying the dataset

```{r modify}
flightData <- mutate(flightData, DAY_OF_WEEK = as.factor(DAY_OF_WEEK))
flightData$DAY_OF_WEEK <- ordered(flightData$DAY_OF_WEEK, levels = c("1","2","3","4","5","6","7"))
levels(flightData$DAY_OF_WEEK) <- c("Sunday","Monday","Tuesday","Wednesday","Thursday","Friday","Saturday")
```

####Additional modification
Modifying the dataset convert integer time to HH:MM format time
```{r}
flightData <- mutate(flightData, SCHEDULED_DEPARTURE = format(strptime(sprintf("%04d",SCHEDULED_DEPARTURE), format="%H%M"), format = "%H:%M"))
flightData <- mutate(flightData, DEPARTURE_TIME = format(strptime(sprintf("%04d",DEPARTURE_TIME), format="%H%M"), format = "%H:%M"))
```

Modifying the dataset to have, a column for morning to night clasification for _scheduled_ departure time
```{r}
flightData <- mutate(flightData, DEP_TIME_OF_DAY = 
                       ifelse (SCHEDULED_DEPARTURE >= "18:00","Evening",
                       ifelse (SCHEDULED_DEPARTURE >= "12:00","Afternoon",
                       ifelse (SCHEDULED_DEPARTURE >= "06:00","Morning","Night")))
                    )
```

Ordering the factors in Time of day
```{r}
flightData <- mutate(flightData, DEP_TIME_OF_DAY = as.factor(DEP_TIME_OF_DAY))
flightData$DEP_TIME_OF_DAY <- ordered(flightData$DEP_TIME_OF_DAY, levels = c("Morning","Afternoon", "Evening","Night"))
```
#EXPLORATORY DATA ANALYSIS
This section should have more graphs, charts and bits 'n pieces of other small studies. More so to make the report and presentation look 'fuller'. In my next version of this doc, I will add more. But feel free to suggest.
####Scatterplot of arrival delay vs time
x-axix each day is from 00:00 to 23:59. But I cant figure out a way to display the time without nasty overlapping so I just hid it.
```{r}
scat.plot <- ggplot(data = drop_na(flightData), aes(x = DEPARTURE_TIME, y = ARRIVAL_DELAY)) + geom_point() + facet_grid(cols = vars(DAY_OF_WEEK))
scat.plot + theme(axis.text.x=element_blank())
```

####Plot of arrival delay vs Day of the week + Time of the day
```{r}
groupByDayTime <- group_by(select(flightData, DAY_OF_WEEK, DEP_TIME_OF_DAY, ARRIVAL_DELAY),DAY_OF_WEEK,DEP_TIME_OF_DAY)
groupByDayTime <- groupByDayTime[complete.cases(groupByDayTime),]
delayByDayTime <- summarise(groupByDayTime, AVG_ARR_DELAY = mean(ARRIVAL_DELAY))
ggplot(data = delayByDayTime, aes(x = DEP_TIME_OF_DAY, y = AVG_ARR_DELAY, fill = DAY_OF_WEEK)) + geom_bar(stat = "identity") + guides(fill = FALSE) + facet_wrap(vars(DAY_OF_WEEK),nrow = 1) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

####Correlation plot of all the time based variables
Modification to subset data into a data frame for those time realted columns + Distance.
Then, correlation, then plot.
```{r}
flightDataTime <- select(flightData, DEPARTURE_DELAY:ARRIVAL_DELAY)
flightDataTime.corr <- round(cor(flightDataTime, use = "complete.obs"),2)
flightDataTime.corr
corrplot(flightDataTime.corr, method = "circle")
```

#INFERENCE

This is to prove that for flights departing on saturday evenings, there is definitely a delay on arrival and it is not a result of sample variance.

####Hypothesis Test
H0 : Mean arrival delay on Sat-Evening <= 0
H1 : Mean arival delay on Sat-Evening > 0

Putting saturday arrival delays in  a vector and then running the above hypothesis tests
```{r hypotest1}
arrdelaySat <- filter(flightData, DAY_OF_WEEK == "Saturday", DEP_TIME_OF_DAY == "Evening")$ARRIVAL_DELAY
summary(arrdelaySat)
test_res <- t.test(arrdelaySat, mu = 0, alternative = "greater")
test_res
```
####ANOVA on DAY + TIME of DAY
```{r}
flightData.aov <- mutate(flightData, DAY_TIME = paste(DAY_OF_WEEK,"_",DEP_TIME_OF_DAY))
res.flightData.aov <- aov(ARRIVAL_DELAY ~ DAY_TIME, data = flightData.aov)
summary(res.flightData.aov)
```

#THE 'BEST' MODEL

In the last lecture class, the professor taught us a workflow for determoining the 'best' model. I have used the 'backward elimination adjusted r squared' method. Please use the lecture notes to understand what I did below:

####STEP 1: FULL MODEL (day,time, airline, origin, destination)
```{r step1}
arrDelayModel00 <- lm(ARRIVAL_DELAY ~ DAY_OF_WEEK + DEP_TIME_OF_DAY + AIRLINE + ORIGIN_AIRPORT + DESTINATION_AIRPORT , data = flightData)
summary(arrDelayModel00)$adj.r.squared
```

####STEP 2: 4-Variable MODEL
From the results below it looks like removing any of the variables gives a lower adjusted r-square.
So, **the full model is the best model**
```{r step2, eval = TRUE}
arrDelayModel01 <- lm(ARRIVAL_DELAY ~ DEP_TIME_OF_DAY + AIRLINE + ORIGIN_AIRPORT + DESTINATION_AIRPORT , data = flightData)
print("No DAY_OF_WEEK")
summary(arrDelayModel01)$adj.r.squared
rm(arrDelayModel01)
arrDelayModel02 <- lm(ARRIVAL_DELAY ~ DAY_OF_WEEK + AIRLINE + ORIGIN_AIRPORT + DESTINATION_AIRPORT , data = flightData)
print("No DEP_TIME_OF_DAY")
summary(arrDelayModel02)$adj.r.squared
rm(arrDelayModel02)
arrDelayModel03 <- lm(ARRIVAL_DELAY ~ DAY_OF_WEEK + DEP_TIME_OF_DAY + ORIGIN_AIRPORT + DESTINATION_AIRPORT , data = flightData)
print("No AIRLINE")
summary(arrDelayModel03)$adj.r.squared
rm(arrDelayModel03)
arrDelayModel04 <- lm(ARRIVAL_DELAY ~ DAY_OF_WEEK + DEP_TIME_OF_DAY + AIRLINE + DESTINATION_AIRPORT , data = flightData)
print("No ORIGIN_AIRPORT")
summary(arrDelayModel04)$adj.r.squared
rm(arrDelayModel04)
arrDelayModel05 <- lm(ARRIVAL_DELAY ~ DAY_OF_WEEK + DEP_TIME_OF_DAY + AIRLINE + ORIGIN_AIRPORT, data = flightData)
print("No DESTINATION_AIRPORT")
summary(arrDelayModel05)$adj.r.squared
rm(arrDelayModel05)
```
#PREDICTION

#### Setting prediction Model for day, time, airline, origin & destination (the winning model above)
```{r}
arrDelayModel <- arrDelayModel00
summary(arrDelayModel)
```

#### Predicting for a specific query to demonstrate prediction power
Predicting ARRIVAL_DELAY for a flight departing from Oakland, on a friday afternoon for Baltimore via Spirit Airlines
```{r}
predict(arrDelayModel, data.frame(DAY_OF_WEEK = "Friday", DEP_TIME_OF_DAY = "Afternoon", AIRLINE = "NK", ORIGIN_AIRPORT = "OAK", DESTINATION_AIRPORT = "BWI"))
```
#APPENDIX (All extra bits of analysis that were superceded by the above code)
Plot of arrival delay vs Day of the week

```{r}
groupByDay <- group_by(select(flightData, DAY_OF_WEEK, ARRIVAL_DELAY),DAY_OF_WEEK)
groupByDay <- groupByDay[complete.cases(groupByDay),]
delayByDay <- summarise(groupByDay, AVG_ARR_DELAY = mean(ARRIVAL_DELAY))
ggplot(data = delayByDay, aes(x = DAY_OF_WEEK, y = AVG_ARR_DELAY, fill = DAY_OF_WEEK)) + geom_bar(stat = "identity") + guides(fill = FALSE)
```

####Hypothesis Test
H0 : Mean arrival delay on Sat <= 0
H1 : Mean arival delay > 0

Putting saturday arrival delays in  a vector and then running the above hypothesis tests
```{r hypotest2}
arrdelaySat <- filter(flightData, DAY_OF_WEEK == "Saturday")$ARRIVAL_DELAY
summary(arrdelaySat)
test_res <- t.test(arrdelaySat, mu = 0, alternative = "greater")
test_res
```
####Anova on day of the week
```{r}
res.aov <- aov(ARRIVAL_DELAY ~ DAY_OF_WEEK, data = flightData)
summary(res.aov)
```